Quantifying and Modeling Global Water Accessibility including Water Typology

### 2021-01-27 docs/manuscript/manuscript.tex
Simplified contribution section for chronology.

\section{Contribution}
Lack of water accessibility is a major crisis for many vulnerable people around the world. We have determined to seek literature reviews from the past 15 years to have the most up to date information because information on water accessibility is a topic that is constantly changing throughout the world due to different international, government, and even local policies. The literature review intends to investigate different journal articles on water accessibility in a global scale. Therefore we can conduct a research to fill many major gaps  to analyze the water accessibility predictor variables which could increase and enhance water accessibility in global scale. We will always gain new insight and wisdom on information about water accessibility through the global scale as we investigate and determine literature reviews chronologically. 
\subsection{2006-2016}
Improving water accessibility have been a priority by many organizations around the world, realizing it is the basic human rights to have clean and accessible drinkable water in past 15 years. During 2006 to 2016, overall water accessibility has been improving by utilizing the indicators of the past \citep{fullerTrackingProgressGlobal2016}. However, some researchers raises concerns on lack of clarity with the global water accessibility monitoring \citep{ondaGlobalAccessSafe2012, jiaHighResolutionSpatialDistribution2016}. However, in an era of rapidly changing advanced technologies, international, and government policies, some of the older research might not be adequate for up to date information for  example of The eight Millennium Development Goals (MDGs) upgrading into The Sustainable Development Goals (SDGs). In light of the newest SDGs of providing clean and accessible drinking water for all by 2030, researchers have devoted their study to understand water accessibility in global and specific country levels. Several research and studies have focused global water accessibility ,and its progress \citep{sorensonSafeAccessSafe2011,bainAccountingWaterQuality2012,wssccWaterSanitationHygiene2014,  ondaGlobalAccessSafe2012, bartramGlobalMonitoringWater2014, ondaCountryClusteringApplied2014, whoDrinkingwater,fullerTrackingProgressGlobal2016, weiyuyuGlobalPerspectiveDrinkingwater2016, grahamAnalysisWaterCollection2016,overboOnplotDrinkingWater2016}. While there has been studies focusing on local community in different parts of the country learning about their water accessibility and sanitation initiation \citep{jagalsDoesImprovedAccess2006, geereDomesticWaterCarrying2010, simaWaterFlowsEnergy2013, hoChallengeGlobalWater2014, sakisakaIntroductionCommunityWater2015, jiaHighResolutionSpatialDistribution2016}. Where majority raised concerns about lack of monitoring sustainable water accessibility measurements for many people, and even few argues for greater focus in many people's access to drinking water sources. Nonetheless, there is a lack of robust study on modes of transportation related to enhance water accessibility for all people by 2030.

\citet{jagalsDoesImprovedAccess2006} highlights in a case study in a rural area in South Africa discovered providing small communities. The rural area used untreated river water as their water source. The Researchers installed good quality water with a piped distribution system and increase accessibility by installing communal taps. However, this implementation did not fall within the parameters of safe water. This didn't increase household quantities, which remained around 18 L per person per day. Their research
gives a potential understanding even making the drinking water accessible might not improve the quality of the water
itself.

\citet{ondaGlobalAccessSafe2012} discusses the global access to safe water by accounting for water quality and the resulting
impact on (MDGs) progress. They altered the current Joint Monitoring Programme (JMP) estimation by accounting for microbial
water quality and sanitary risk utilizing national water quality data. 
Overall, their analysis highlights the potential fatigue can have causes that aren't due to underlying disease. The main example would be heavy exertion of carrying water.  Improvement of modes of transportation can enhance water accessibility which can improve one's health overall. 
modes of transportation can affect one's health.

\citet{simaWaterFlowsEnergy2013} highlights topic where it discovered that over 76 percent of the Kenya's city's water
is utilized by less than ten percent of household piped water.  The study is focused on Kisumu Kenya and Sima
interviewed 260 informal water business operators. This article revealed a new insight where the majority of the
population in Kisumu relied on purchased water from kiosks (1.5 million $m^{3}$ per day) and used hand-drawn water-carts
(0.75 million $m^{3}$ per day)\citep{simaWaterFlowsEnergy2013}. Also, we discovered that the water trucking industry utilized the
most energy consumption and was high cost: water delivery trucks have the highest per cubic meter energy demand (35
$MJ/m^{3}$)\citep{simaWaterFlowsEnergy2013}.

\citet{hoChallengeGlobalWater2014} investigates a head-to-head comparison of such indicators with other possible distance and time
metrics among rural 1,103 households in Nampula province, Mozambique. The researchers found out that Euclidean line
distance to be a good representative for route distance ($R^{2} = 0.98$), while self-reported travel time is a poor
representative ($R^{2} = 0.12$). One of the key insights was that a 15-minute decrease in one-way walk time to a water source
is related to a 41\% reduction in diarrhea, improved children's nutrition, and an 11\% reduction in under-five child
mortality rate \citep{hoChallengeGlobalWater2014}.

\citet{bartramGlobalMonitoringWater2014} highlights background information about the Millennium Development Goals. How it has
established global targets for drinking water and sanitation access. The article directly gives
much information about progress towards focused targets, directed by international monitoring, has reduced the global
disease burden and improved the quality of life \citep{bartramGlobalMonitoringWater2014}. It utilizes a variety of sources such as
JMP, DHS, MICS surveys to analyze the data of many different locations in the world, therefore, it is a broad scope of
study of the progress of accessible and drinking water and sanitation globally. Our research can study further what
modes of transportation of clean drinking water would affect the progress of improving drinking water and sanitation in
one's community.

\citet{ondaCountryClusteringApplied2014} proposed a country typology based on nine water and sanitation indicators in a global scale. They applied hierarchical clustering to 156 countries, accounting for 6.75 billion people (93\%). This work improves on previous geography or income-based country clustering.

\citet{whoDrinkingwater} article gives an ambitious plan to reach universal drinking water by the year 2030 that is
affordable and safe followed by the Sustainable Development Goal six goals. The WHO/UNICEF Joint Monitoring Program for
Water Supply and Sanitation (JMP) have been globally monitoring progress on drinking water and sanitation since 1990. \citet{whoDrinkingwater} outlines different plans with consistent traits to determine what is deemed to be safe and affordable drinking
water and good sanitation.

\citet{fullerTrackingProgressGlobal2016} analyzes the global accessibility to safe drinking water and sanitation and concludes that
it has been improving well during the Millennium Development Goal Period. \citet{fullerTrackingProgressGlobal2016} chose some countries and studied the overall countries
progress which in his methodology revealed sigmoidal trends. Our study can reveal if certain methods of transportation
of drinking water and sanitation may have influenced the effectiveness of the improved drinking water met by MDGs
standards.

\citet{jiaHighResolutionSpatialDistribution2016} highlights that there is inequality within safe sanitation. There is a great difference in
wealth levels in many low-income countries. Therefore, this will hinder the improvement of MDG
progress. \citet{jiaHighResolutionSpatialDistribution2016}discusses future interventions to be engaged with different wealth
categories. Our study can assist in understanding improved sanitation increase by making inferences about how they
receive water may give the community focusing on sanitation. Overall, income inequality affects who will have access to
proper sanitation.

\subsection{2017-Present}
\citet{cassiviAccessDrinkingWater2018} highlights about how there should be more information to the effect of including a 30-minute
collection time to monitor access to drinking water. The lack of access to on-premises water sources can cause the use
of greater distant alternative sources. Therefore causing the quantity of retrieved water to reduce as the water plateau
phenomenon.\citet{cassiviAccessDrinkingWater2018} hypothesizes that the households who must travel further are using unimproved
sources. For our research, we can discover what modes of transportation people are traveling to retrieve the water and
in what quantity depending on their means of transportation. Therefore, learning how to decrease the travel time of
people to retrieve the water because according to the article, it can be a great burden and does not increase the
quality of one's life to retrieve water daily and far away from their homes.

>>>>

\section{Contribution}
Lack of water accessibility is a major crisis for many vulnerable people around the world. We have determined to seek literature reviews from the past 15 years to have the most up to date information because information on water accessibility is a topic that is constantly changing throughout the world due to different international, government, and even local policies. The literature review intends to investigate different journal articles on water accessibility in a global scale. Therefore we can conduct a research to fill many major gaps  to analyze the water accessibility predictor variables which could increase and enhance water accessibility in global scale. We will always gain new insight and wisdom on information about water accessibility through the global scale as we investigate and determine literature reviews chronologically. 
\subsection{2006-2016}
Improving water accessibility have been a priority by many organizations around the world, realizing it is the basic human rights to have clean and accessible drinkable water in past 15 years. During 2006 to 2016, overall water accessibility has been improving by utilizing the indicators of the past \citep{fullerTrackingProgressGlobal2016}. However, some researchers raises concerns on lack of clarity with the global water accessibility monitoring \citep{ondaGlobalAccessSafe2012, jiaHighResolutionSpatialDistribution2016}. However, in an era of rapidly changing advanced technologies, international, and government policies, some of the older research might not be adequate for up to date information for  example of The eight Millennium Development Goals (MDGs) upgrading into The Sustainable Development Goals (SDGs). In light of the newest SDGs of providing clean and accessible drinking water for all by 2030, researchers have devoted their study to understand water accessibility in global and specific country levels. Several research and studies have focused global water accessibility ,and its progress \citep{sorensonSafeAccessSafe2011,bainAccountingWaterQuality2012,wssccWaterSanitationHygiene2014,  ondaGlobalAccessSafe2012, bartramGlobalMonitoringWater2014, ondaCountryClusteringApplied2014, whoDrinkingwater,fullerTrackingProgressGlobal2016, weiyuyuGlobalPerspectiveDrinkingwater2016, grahamAnalysisWaterCollection2016,overboOnplotDrinkingWater2016}. While there has been studies focusing on local community in different parts of the country learning about their water accessibility and sanitation initiation \citep{jagalsDoesImprovedAccess2006, geereDomesticWaterCarrying2010, simaWaterFlowsEnergy2013, hoChallengeGlobalWater2014, sakisakaIntroductionCommunityWater2015, jiaHighResolutionSpatialDistribution2016}. Where majority raised concerns about lack of monitoring sustainable water accessibility measurements for many people, and even few argues for greater focus in many people's access to drinking water sources. Nonetheless, there is a lack of robust study on modes of transportation related to enhance water accessibility for all people by 2030. 
\subsection{2017-Present}
\citet{cassiviAccessDrinkingWater2018} highlights about how there should be more information to the effect of including a 30-minute
collection time to monitor access to drinking water. The lack of access to on-premises water sources can cause the use
of greater distant alternative sources. Therefore causing the quantity of retrieved water to reduce as the water plateau
phenomenon.\citet{cassiviAccessDrinkingWater2018} hypothesizes that the households who must travel further are using unimproved
sources. For our research, we can discover what modes of transportation people are traveling to retrieve the water and
in what quantity depending on their means of transportation. Therefore, learning how to decrease the travel time of
people to retrieve the water because according to the article, it can be a great burden and does not increase the
quality of one's life to retrieve water daily and far away from their homes.
